---
title: |
  | Supporting Information for:
  | Comparison of chromatographic stationary phases using a Bayesian-based multilevel model
author:
  - name: "Paweł Wiczling*"
    affiliations:
      - name: "Department of Biopharmaceutics and Pharmacodynamics, Medical University of Gdańsk, Gen. J. Hallera 107, 80-416 Gdańsk, Poland"
  - name: "Agnieszka Kamedulska"
    affiliations:
      - name: "Department of Biopharmaceutics and Pharmacodynamics, Medical University of Gdańsk, Gen. J. Hallera 107, 80-416 Gdańsk, Poland"
date: "`r format(Sys.Date())`"
format:
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
    fig-width: 7
    fig-height: 7
    keep-tex: true
    template-partials: 
      - title.tex
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE, message=FALSE, error=FALSE, warning=FALSE, echo=FALSE, comment=NA, out.width='95%')
```

```{r, include=FALSE}
library(dplyr)
library(kableExtra)
library(cmdstanr)
```


{{< pagebreak >}}

# Table S1. Summary of the MCMC simulations of the marginal posterior distributions of population-level model parameters. 

Mean denotes sample mean, MCSE denotes Monte Carlo Standard Error, StdDev denotes sample standard deviation, 5%, 50%, 95% denote corresponding quantiles, N_Eff denotes effective sample size, R_Hat denotes a measure of chain equilibrium must be within 0.05 of 1.0.

```{r}
fit <- cmdstanr::as_cmdstan_fit(c(
                                  '../stanfiles/output_1.csv',
                                  '../stanfiles/output_2.csv',
                                  '../stanfiles/output_3.csv',
                                  '../stanfiles/output_4.csv',
                                  '../stanfiles/output_5.csv',
                                  '../stanfiles/output_6.csv',
                                  '../stanfiles/output_7.csv',
                                  '../stanfiles/output_8.csv'
                                ))
fit$print(c("logkwHat"), max_rows = 26)
```

{{< pagebreak >}}

# Figure S1. Raw data for 6 selected analytes.

![](../figures/rawdata/Acridine.pdf)
{{< pagebreak >}}
![](../figures/rawdata/Baclofen.pdf)

{{< pagebreak >}}
![](../figures/rawdata/Hydrocortisone.pdf)

{{< pagebreak >}}
![](../figures/rawdata/Pioglitazone.pdf)
{{< pagebreak >}}
![](../figures/rawdata/Quinine.pdf)

{{< pagebreak >}}
![](../figures/rawdata/Tolbutamide.pdf)

{{< pagebreak >}}



# Licenses {.unnumbered}

-   Code & copy; 2023, Paweł Wiczling, licensed under BSD-3.
-   Text & copy; 2023, Paweł Wiczling, licensed under CC-BY-NC 4.0.

# Original Computing Environment {.unnumbered}

```{r}
sessionInfo()
```
