---
title: "QSRR revisited. Hierarchical models"
author:
  - name: "Paweł Wiczling*"
    affiliations:
      - name: "Department of Biopharmaceutics and Pharmacodynamics, Medical University of Gdańsk, Gen. J. Hallera 107, 80-416 Gdańsk, Poland"
date: "`r format(Sys.Date())`"
format:
  html:
    theme: cosmo
    toc: true
    code-fold: true  
    code-tools: true
    fig-width: 7
    fig-height: 7
knitr:
  opts_chunk: 
    dev: "ragg_png"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=FALSE, message=FALSE, error=FALSE, warning=FALSE, comment=NA, out.width='95%')
```

# Introductions

...

# Setup
```{r message=FALSE}
library(bbr)
library(bbr.bayes)
library(dplyr)
library(ggplot2)
library(patchwork)
library(gridExtra)
library(naniar)
library(knitr)
library(data.table)
library(tidyverse)
library(glue)
library(whisker)
library(here)
library(cmdstanr)
library(gridExtra)
library(posterior)
library(bayesplot)
library(tidybayes)
library(reshape2)
library(pracma)
library(mrgmisc)
#remotes::install_github("metrumresearchgroup/mrgmisc")
set_cmdstan_path("C:/Users/GUMed/.cmdstan/cmdstan-2.36.0")
set.seed(10271998) ## not required but assures repeatable results
```

```{r settings}
data_dir = here::here("data")
model_dir <- here::here("model","stan")  
figures_dir <- here::here("deliv","figures", "stan")
tables_dir <- here::here("deliv","tables", "stan")
if(!file.exists(figures_dir)) dir.create(figures_dir, recursive = T)
if(!file.exists(model_dir)) dir.create(model_dir, recursive = T)
if(!file.exists(tables_dir)) dir.create(tables_dir, recursive = T)
```


We used a publicly available [dataset](www.retentionprediction.org/hplc/database/) that comprises the measurements of RP-HPLC retention times collected for 1026 analytes. The retention times were measured under isocratic conditions on Eclipse Plus C18 (Agilent) stationary phase with 3.5 μm particles. The experiments were conducted using a mixture of two solvents: solvent A, which was made of 0.1% formic acid in water, and solvent B, which was made of 0.1% formic acid in acetonitrile. The column temperature was set at 35\^{\circ}C. The data were collected by Boswell et al. and were used to create a method to predict retention time by Back-Calculating the Gradient.

The values of molecular mass MM were added to the dataset. They were calculated using [ACD/Labs program](www.acdlabs.com) based on the provided structures of analytes. The MM ranged from 73.09 to 656.8. The functional groups present in the analyzed chemical compounds are presented below, along with their frequencies


```{r load-data, message=FALSE, warning=FALSE}
# load data
data       <- read.csv("../data/database_logk_1026.csv", header = TRUE)
analytes_names  <- read.csv("../data/database_logk_1026_analyte_names.csv", header = TRUE)

data<-data %>%left_join(analytes_names) 

functional_groups = read.csv('../data/checkmol_functional_groups.csv')
functional_groups_names = read.csv('../data/checkmol_functional_group_names.csv')

# combine nr of caroboxylic acid and carboxyalic acid salt functional groups
# heterocyclic compounds with more than 6 heterocycles are treated as if they have six
functional_groups[,76]=functional_groups[,76]+functional_groups[,77]       
functional_groups[which(functional_groups[,202]>5.5),202] = 6; 

# exclude some functional groups
idx_excluded <- c(1, 2, 3, 6, 27, 28, 37, 47, 48, 51, 55, 61, 62, 67, 73, 74, 75, 77, 80, 91, 99, 109, 116, 117, 121, 125, 129, 142, 153, 154, 160, 161, 168, 173, 178, 181, 182, 186, 187, 191, 196, 201, 202)

functional_groups_names <- functional_groups_names[-idx_excluded,]
functional_groups <- functional_groups[,-idx_excluded]

rm(idx_included)

totalnrgroups <- summarise_each(functional_groups, funs(sum))

# remove functional groups not present in the dataset:
functional_groups <- functional_groups[,which(totalnrgroups!=0)]
functional_groups_names<- functional_groups_names[which(totalnrgroups!=0),]
totalnrgroups <- summarise_each(functional_groups, funs(sum))

data_ACD = read.csv('../data/ACD_pKas.csv')
data_ACD$R = rowSums(data_ACD[,3:5]<14) # No of dissociation steps
```

# Prepare data for calculations

```{r}
df <- data %>% distinct(ID, .keep_all = TRUE) 
pKaslit = data_ACD[,3:5]           # pKa values as predicted by ACD
pKasliterror = data_ACD[,17:19]    # pKa error as predicted by ACD
chargesA = abs(data_ACD[,9:12])    # number of ionized groups (anions)
chargesB = abs(data_ACD[,13:16])   # number of ionized groups (cations)
charges = chargesA+chargesB                       # absolute charge
groupsA = (chargesA[,2:4]-chargesA[,1:3])         # acidic group
groupsB = -(chargesB[,2:4]-chargesB[,1:3])        # basic group
R = data_ACD$R                                    # number of dissociation steps
groups = groupsB-groupsA
logPobs =df$logP_ACD 
logPobs[526] = mean(logPobs, na.rm = TRUE)

# identify acidic groups
idxGroupsA <-which(groupsA!=0,arr.ind = T)
nGroupsA <- nrow(idxGroupsA)
pKaslitA<-pKaslit[idxGroupsA]

# identify basic groups
idxGroupsB <-which(groupsB!=0,arr.ind = T)
nGroupsB <- nrow(idxGroupsB)
pKaslitB<-pKaslit[idxGroupsB]

# groups dissociated in the whole pH range
chargesA0 <- chargesA[,1]
chargesB0 <- chargesB[,max(R)]

idxA0 = c(which(chargesA0==1), which(chargesA0==2))
idxB0 = c(which(chargesB0==1), which(chargesB0==2))

nA0 <- length(idxA0)
nB0 <- length(idxB0)

nObs <- length(data$ID)
nAnalytes <- length(unique(data$ID))

start <- (1:nObs)[!duplicated(data$ID)]
end <- c(start[-1] - 1, nObs)
```

```{r initial-estimates}

initial_param <- matrix(NA,nAnalytes,3)

for(i in 1:nAnalytes){
  if(length(data$fi[data$ID==i])==1){
    initial_param[i,2] <- polyfit(-16*data$fi[data$ID==i]/(1+2*data$fi[data$ID==i]), data$logk[data$ID==i],0)
    initial_param[i,1] <- -16
  }else{
    initial_param[i,1:2] <- polyfit(data$fi[data$ID==i]/(1+2*data$fi[data$ID==i]), data$logk[data$ID==i],1)
  }
}

initial_param[,3] <- rep(2,nAnalytes)

init_aprox <-  data.frame(logkw = initial_param[,2])
init_aprox$S2 <- rep(2,nAnalytes)
init_aprox$S1 <- -initial_param[,1]/(1+2)
init_aprox$chargesB0 <-chargesB[,2]
init_aprox$chargesA0 <-chargesA0
```


# Data analysis

## Model

The data was analyzed using a ...

$$
\begin{aligned}
& logk_{i,j}  = logkw_{i} - S_{1,i} \cdot (1 + S_{2}) \cdot \varphi_{i,j} / (1 + S_{2} \cdot \varphi_{i,j})
\end{aligned}
$$
The statistical model has the following structure (hierarchical model):

$$ 
\begin{aligned}
& \log k_{obs_{i,j}} \sim student_t(\nu,logk_{i,j},\sigma) \\
& R_{i} \sim \text{MVN}(\theta_R \cdot fr_i + \theta_{dR} \cdot (1-fr_i)  + \beta \cdot log P_i , \Omega) \\
\end{aligned}
$$
Needs improvements!! where $R_i=(logkw_{i}, S_{1,i})$ is a vector of subject-specific parameters, $logk_{i,j}$ corresponds to the above Neue equation, $MVN$ is a multivariate normal distribution, $\theta_P$ is a vector of typical values of $R_i$, $beta_R$ is a vector of slopes between $R_i$ and $logP_i$. In turn, $\sigma$ is the scale and $\Omega$ is the scale matrix for the random effects related to the residual and unexplained between-analyte variability. $fr$ is a fraction of dissocation $\Omega$ was decomposed to:

$$
\Omega = diag(\omega) \cdot \rho \cdot diag(\omega) 
$$

where $\rho$ denotes correlation a correlation matrix and $\omega$ denotes standard deviation for between analyte variability.

## Priors

The priors were specified as follows:

$$ 
\begin{aligned}
& \theta_{logkw} \sim normal(2, 4), \\
& \theta_{S1} \sim normal(4, 2), \\
& \theta_{S2} \sim lognormal(0.693, 0.125), \\
& \theta_{dlogkw} \sim normal(-1, 0.25), \\
& \theta_{dS1} \sim normal(0, 0.5), \\
& \nu = 7,\\
& \omega_{logkw},\omega_{S1} \sim normal_+(0, 1), \\
& \beta_{logkw} \sim  normal(0.7, 0.125), \\
& \beta_{S1} \sim  normal(0.5, 0.125), \\
& \sigma \sim normal_+(0,0.05), \\
& p(\rho) \propto LKJ(2) \cdot \Pi_u N(c_u, 0.125), \\
& c_u = [0.75]
\end{aligned}
$$

LKJ(2) ensure that the density is uniform over correlation matrices of order 2 and u denotes the unique lower triangular elements of correlation matrix (<http://srmart.in/informative-priors-for-correlation-matrices-an-easy-approach/>)

# Stan

Multilevel modeling was performed in [Stan software](https://mc-stan.org/) linked with R/ [cmdstanr](https://mc-stan.org/cmdstanr/). For the inference we used 500 iterations, 1000 warmup iterations, and 8 Markov chains. The reduce_sum function was selected to accelerate the calculations. It works by parallelizing the execution of a single Stan chain across multiple cores. Convergence diagnostics were checked using Gelman−Rubin statistics and trace plots.

### Initialize variables and parameters

```{r stansetup}
# create Stan data set:
datastruct <- with(data,
                  list(nAnalytes=length(unique(data$ID)),
                       nObs=length(data$ID),
                       analyte=match(data$ID, unique(data$ID)),
                       fi=data$fi,
                       logPobs=logPobs, 
                       start = start,
                       end= end,
                       pKaslit = pKaslit,
                       charges = charges,
                       logkobs=logk,
                       run_estimation=1))

# initialize the values for each variable in each chain:
init <- function(){
  list(   logkwHat  = rnorm(1,2,2),
          S1Hat    = rnorm(1,5,1),
          dlogkwHat  = rnorm(1,-1,0.125),
          dS1Hat    = rnorm(1,0,0.5),
          S2Hat = 2*exp(rnorm(1,0,0.125)),
          beta  = rnorm(2,c(0.7,0.5),c(0.125,0.5)),
          omega = c(1,1)*exp(rnorm(2, 0, 0.5)),
          rho = matrix(c(1,0.75,0.75,1), nrow=2),
          #param = cbind(matrix(2, 1, nrow=nSteps), matrix(5, 1, nrow=nSteps)),
          param = cbind(init_aprox$logkw,init_aprox$S1),
          sigma  =  0.05*exp(rnorm(1, 0, 0.5))
  )
}

```

### Fitting the model
We compiled the model using cmdstanr:
```{r mod-settings}
model_name <- "mod1"
model_name_ext <- paste0(model_name,".stan")
#' folder shortcuts:
mod_figures_dir <- here::here("deliv", "figures", model_name)
mod_tabels_dir  <- here::here("deliv", "tables", model_name)
mod_model_dir   <- here::here("model", "stan", model_name)
mod_output_dir   <- here::here("model", "stan", model_name, "output")

if (!file.exists(mod_figures_dir)) dir.create(mod_figures_dir)
if (!file.exists(mod_tabels_dir)) dir.create(mod_tabels_dir)
if (!file.exists(mod_model_dir)) dir.create(mod_model_dir)
if (!file.exists(mod_output_dir)) dir.create(mod_output_dir)
```

```{r stancompile, eval = FALSE}
mod1 <- cmdstan_model(here::here(mod_model_dir, model_name_ext), stanc_options = list("O1"))
```

We used optimization for initial testing:

```{r stanoptimize, eval = FALSE}
fit_opt <- mod1$optimize(
  data = datastruct,
  output_dir = mod_output_dir,
  output_basename = "mod1-opt",
  init = init,
  iter=1000
)

fit_opt$print(max_rows=100)
```

```{r optplots, eval=FALSE}
data$logkopt<-fit_opt$mle("logkx")

datax<-data %>%
  filter(ID %in% c(20:30)) %>%
  ggplot(aes(x=fi, y=logk)) + 
  geom_point() + 
  geom_line(aes(x=fi, y=logkopt)) + 
  facet_wrap(~ID) +
  labs(x = "\u03C6", y = expression(logk)) + 
  theme_gray(base_size = 14) + 
  theme(legend.position="none")
datax
```

For local computations one can use cmdstanr:

```{r stansample, eval = FALSE}
 fit <- mod1$sample(
   data = datastruct,
   output_dir = mod_output_dir,
   output_basename = "mod1-loc",
   init = init,
   iter_warmup = 500,
   iter_sampling = 250,
   chains = 4,
   parallel_chains = 4,
   refresh = 100,
   adapt_delta=0.9
 )

# fit$save_object(file = here(mod_output_dir,"mod1-loc.RDS"))
```

We performed out computations at the Academic Computer Center in Gdańsk, [Tryton Cluster](https://docs.task.gda.pl/kdm/zasoby-sprzetowe/tryton/). In this case:

1.  we dumped the necessary data to .json format

```{r dumpfiles, eval = FALSE}
write_stan_json(datastruct, paste0(mod_model_dir,"/standata.json"), always_decimal = FALSE)
write_stan_json(init(), paste0(mod_model_dir,"/init-1.json"), always_decimal = FALSE)
write_stan_json(init(), paste0(mod_model_dir,"/init-2.json"), always_decimal = FALSE)
write_stan_json(init(), paste0(mod_model_dir,"/init-3.json"), always_decimal = FALSE)
write_stan_json(init(), paste0(mod_model_dir,"/init-4.json"), always_decimal = FALSE)
write_stan_json(init(), paste0(mod_model_dir,"/init-5.json"), always_decimal = FALSE)
write_stan_json(init(), paste0(mod_model_dir,"/init-6.json"), always_decimal = FALSE)
write_stan_json(init(), paste0(mod_model_dir,"/init-7.json"), always_decimal = FALSE)
write_stan_json(init(), paste0(mod_model_dir,"/init-8.json"), always_decimal = FALSE)
```

2.  we run the model using the batch file:

```{=html}
<pre> 
</pre>
```

3.  After calculations, we loaded the output files using cmdstanr. 

```{r stanload}
fit <- cmdstanr::as_cmdstan_fit(c(
                                  paste0(mod_output_dir,"/",  model_name, "-1.csv"),
                                  paste0(mod_output_dir,"/",  model_name, "-2.csv"),
                                  paste0(mod_output_dir,"/",  model_name, "-3.csv"),
                                  paste0(mod_output_dir,"/",  model_name, "-4.csv"),
                                  paste0(mod_output_dir,"/",  model_name, "-5.csv"),
                                  paste0(mod_output_dir,"/",  model_name, "-6.csv"),
                                  paste0(mod_output_dir,"/",  model_name, "-7.csv"),
                                  paste0(mod_output_dir,"/",  model_name, "-8.csv")
                                ))
```

4.  finally we checked the diagnostics of Monte Carlo inferences based on the Stan documentation described herein [diagnose](https://mc-stan.org/docs/cmdstan-guide/diagnose.html)

```{r standiagnose, eval = FALSE}
 #fit$cmdstan_diagnose()
 setwd(mod_output_dir)
 str = paste0(cmdstan_path(), '/bin/diagnose  mod1-*.csv')
 system(str,intern=TRUE)
```

The diagnostics are reasonable given model complexity. Output copied here to save time:

```{=html}
<pre>
</pre>
```

## Summary of model parameters (table):

The [stansummary](https://mc-stan.org/docs/cmdstan-guide/stansummary.html) function was used to report summary and diagnostic statistics over model parameters. 


```{r summary}
fit$print(max_rows = 180)
```


# Genereted quantitites

```{r paths}
modelqg_name <- "mod1gq"
modelqg_name_ext <- paste0(modelqg_name,".stan")
modgq_model_dir   <- here::here("model", "stan", modelqg_name)
modgq_output_dir   <- here::here("model", "stan", modelqg_name, "output")

if (!file.exists(modgq_model_dir)) dir.create(modgq_model_dir)
if (!file.exists(modgq_output_dir)) dir.create(modgq_output_dir)

```

```{r generete-quntities, eval=FALSE}
mod1gq <- cmdstan_model(here(modgq_model_dir,model_name_ext))
fitgq  <- mod1gq$generate_quantities(fit,
                                          data = datastruct,
                                          seed = 123,
                                          parallel_chains = 4,
                                          output_dir = modgq_output_dir,
                                          output_basename = modelqg_name)
```

# Results

## Load
```{r etaload}
x <- cmdstanr::read_cmdstan_csv(c(
                                  paste0(modgq_output_dir, "/",  modelqg_name, "-1.csv"),
                                  paste0(modgq_output_dir, "/",  modelqg_name, "-2.csv"),
                                  paste0(modgq_output_dir, "/",  modelqg_name, "-3.csv"),
                                  paste0(modgq_output_dir, "/",  modelqg_name, "-4.csv")
                                ))
draws_gen_df <- as_draws_df(x$generated_quantities)

```

## Trace plots

Trace plots are time series plots of Markov chains. Here we show standard trace plots for several parameters

```{r trace-plots, include=TRUE}
bayesplot::mcmc_trace(fit$draws(c("logkwHat","S1Hat","dlogkwHat","dS1Hat","beta","S2Hat"))) 
bayesplot::mcmc_trace(fit$draws(c("omega","rho[1,2]","sigma"))) 
bayesplot::mcmc_trace(fit$draws(c("param[626,1]","param[626,2]"))) 
```

## Summary of model parameters (figures)

```{r mcmc-intervals}
mcmc_intervals(fit$draws(c("logkwHat","S1Hat","dlogkwHat","dS1Hat","beta","S2Hat")))
mcmc_intervals(fit$draws(c("omega","rho[1,2]","sigma")))
```

## Fits

```{r ind-pop-gof-plots}

results_ind <-draws_gen_df %>%
  slice_sample(n=200) %>%
  tidybayes::spread_draws(slogk_ind[i]) %>% 
  rename(ID=i) %>%
  group_by(ID)%>%
  tidybayes::median_qi(logkpred = slogk_ind) %>%
  select(-ID)

results_pop <-draws_gen_df %>%
  slice_sample(n=200) %>%
  tidybayes::spread_draws(slogk_pop[i]) %>% 
  rename(ID=i) %>%
  group_by(ID)%>%
  tidybayes::median_qi(logkpred = slogk_pop) %>%
  select(-ID)

p1 = data%>%
  cbind(results_ind) %>%
  ggplot(aes(x=logkpred, y=logk)) + 
  geom_point() + 
  #geom_errorbar(aes(xmin = .lower, xmax = .upper), width=0.01)+
  geom_abline(intercept = 0, slope = 1) +
  labs(x = "logk (individual)", y = "logk (observed)") +
  theme_gray(base_size = 14) + 
  theme(legend.position="none")

p2 = data%>%
  cbind(results_pop) %>%
  ggplot(aes(x=logkpred, y=logk)) + 
  geom_point() + 
  #geom_errorbar(aes(xmin = .lower, xmax = .upper), width=0.01)+
  geom_abline(intercept = 0, slope = 1) +
  labs(x = "logk (population)", y = "logk (observed)") + 
  theme_gray(base_size = 14) + 
  theme(legend.position="none")


p1+p2

```


```{r ind-predictions}

analytes_idx = c(1:9, 501,510, 626)

p <-draws_gen_df %>%
  slice_sample(n=1000) %>%
  tidybayes::spread_draws(sparam_ind[i,..], sS2Hat) %>%
  rename(ID=i) %>%
  left_join(analytes_names) %>%
  filter(ID %in% analytes_idx) %>%
  tidyr::expand_grid(fi = seq(0,1,0.1)) %>%
  mutate(logk = sparam_ind.1-sparam_ind.2*(1+sS2Hat)*fi/(1+sS2Hat*fi)) %>%
  ggplot(aes(x = fi, y = logk)) +
  ggdist::stat_lineribbon(.width = c(.90), alpha = 1/2) +
  geom_point(data = filter(data,ID %in% analytes_idx),aes(x=fi, y=logk)) + 
  facet_wrap(~Analyte)+
  labs(x="\u03C6", y="logk (individual)") +
  theme(legend.position = "none") +
  coord_cartesian(xlim=c(0,1),ylim=c(-1,2))

print(p)
```

```{r pop-predictions}

analytes_idx = c(1:9, 501,510, 626)

p <-draws_gen_df %>%
  slice_sample(n=1000) %>%
  tidybayes::spread_draws(sparam_pop[i,..], sS2Hat) %>%
  rename(ID=i) %>%
  left_join(analytes_names) %>%
  filter(ID %in% analytes_idx) %>%
  tidyr::expand_grid(fi = seq(0,1,0.1)) %>%
  mutate(logk = sparam_pop.1-sparam_pop.2*(1+sS2Hat)*fi/(1+sS2Hat*fi)) %>%
  ggplot(aes(x = fi, y = logk)) +
  ggdist::stat_lineribbon(.width = c(.90), alpha = 1/2) +
  geom_point(data = filter(data,ID %in% analytes_idx),aes(x=fi, y=logk)) + 
  facet_wrap(~Analyte)+
  labs(x="\u03C6", y="logk (population)") +
  theme(legend.position = "none") +
  coord_cartesian(xlim=c(0,1),ylim=c(-1,2))

print(p)
```

## Individual Parameters and etaplots

Individual parameter are the analyte-specific parameters estimated by the model. The following plots allow to assess the correlations between these parameters.

### Individual parameters

```{r iparam-column-neutral}

param <- apply(draws_gen_df[,which(colnames(draws_gen_df) %in% grep("^sparam_ind", 
                                                                          names(draws_gen_df), value = TRUE))], MARGIN = 2, FUN = mean)
param <- melt(param)
param1 <- param[1:nAnalytes,]
param2 <- param[(nAnalytes+1):(2*nAnalytes),]

data_to_plot_param <- cbind(param1,param2,logPobs)
colnames(data_to_plot_param) <- c(expression('logkwN'[i]),expression('S1mN'[i]),expression('logP'[i]))

p<-ggpairs(as.data.frame(data_to_plot_param), columnLabels = colnames(data_to_plot_param) ,
        labeller = "label_parsed",upper = list(continuous = "points"))

print(p)
```

### Etas

```{r ieta}

param <- apply(draws_gen_df[,which(colnames(draws_gen_df) %in% grep("^seta_ind", 
                                                                          names(draws_gen_df), value = TRUE))], MARGIN = 2, FUN = mean)
param <- melt(param)
param1 <- param[1:nAnalytes,]
param2 <- param[(nAnalytes+1):(2*nAnalytes),]

data_to_plot_param <- cbind(param1,param2,logPobs)
colnames(data_to_plot_param) <- c(expression('etalogkwN'[i]),expression('etaS1mN'[i]),expression('logP'[i]))

p<-ggpairs(as.data.frame(data_to_plot_param), columnLabels = colnames(data_to_plot_param) ,
        labeller = "label_parsed",upper = list(continuous = "points"))

print(p)
```

# Session info

```{r}
sessionInfo()
```